Formalizing the Kolmogorov extension Theorem in Lean

### 2023-09-16 Manuscript/proof_outline.tex
minor
manuscript semi-final

\documentclass[sigplan,screen]{acmart}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{ulem}
%\usepackage{amsmath}

@@

\usepackage{minted}
\setminted{encoding=utf-8}
\usepackage{fontspec}
%\setmainfont{dejavu-serif}
\setmonofont[Scale=0.833]{dejavu-sans}
%\setmainfont{FreeSerif}
%\setmonofont{FreeMono}
\usepackage{xcolor, graphics}
\definecolor{mygray}{rgb}{0.97,0.97,0.97}
\definecolor{LightCyan}{rgb}{0.8784,1,1}
\newcommand{\leanline}[1]{\mintinline[breaklines, breakafter=., breaklines]{Lean}{#1}}%\textsuperscript{\textcolor{gray}{\tiny(m)}}}
\newcommand{\leanlinecolor}{\mintinline[breaklines, breakafter=., breaklines, bgcolor=LightCyan]{Lean}}
%\usepackage{refcheck}

@@

\begin{abstract}
  We present a formalization in Lean of the Kolmogorov extension theorem,
  which is a main building block in the construction of the (distribution of)
  stochastic processes with arbitrary index sets. Our
  approach is based on mathlib, the mathematical library for Lean. On
  our way, we provide a formalization of the Caratheodory extension
  theorem, which allows to extend a set function on a semi-ring to a
  proper measure on the $\sigma$-algebra generated by the semi-ring. A
  slight generalization of the classical Kolmogorov extension theorem
  allows us to construct the (distribution of) a stochastic process on
  a complete and separable pseudo-metric space rather than a metric
  space.
\end{abstract}

@@

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. {\color{blue}As an application of our implementation, we construct a probability measure on an infinite product space where all coordinates are independent; see Section~\ref{ss:concl}.}

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:}
While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families.
However, the structure of the argument is similar and could use some of our results, like the proof of $\sigma$-additivity from continuity at the empty set (see Lemma~\ref{P:stetigmass}).
A formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} without requirements on $(\alpha_t)_{t\in\iota}$.
All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

We now come to our contribution. 
The extension theorem is a statement about extending a
set-function on a product space to a (finite) measure, where the
product space can come with an arbitraty index set.  The next
definition covers the important concept of a projective family of
measures. In short, we define measures on any finite subset of indices
in a consistent way. Finite sets (of some type \leanline{ι}) are
formalized with \leanline{Finset ι}. % Such a \leanline{Finset} is
defined as a \leanline{Multiset} (a
%quotient type of the \leanline{List}-type with respect to permutations
%of the list) together with a proof that it contains no duplicates\todo{remove this implementation detail?}.

@@

In contrast to statements already part of
\leanline{mathlib}, in the sequel we \leanlinecolor{highlight}
definitions and statements which are part of our own code. In the
formalization of the above, we use \leanline{variable {ι : Type _} {α: ι → Type _}},
which fixes the index set $\iota$ and all spaces
$\alpha_t, t\in\iota$ as global variables.
In Lean, the product space is then \leanline{∀ j : ι, α j} and the product over a finite set \leanline{J} is \leanline{∀ j : J, α j}.

@@

\noindent
It is worth understanding the precise connection of
\leanlinecolor{isProjective} and
\leanlinecolor{isProjectiveMeasureFamily}. In the latter, the first
variable of \leanlinecolor{isProjective} is the family \leanline{P} of
finite measures for all finite subsets of \leanline{ι}. The second
variable is the functions which maps two sets \leanline{I J : Finset
  ι} and a proof \leanline{hJI} of \leanline{J ⊆ I} together with
\leanline{P I} (which is \leanline{μ} in the statement) to the image
measure on \leanline{J}, which is \leanline{μ.map (fun x : ∀ i : I, α
  i => fun j => x ⟨j, hJI j.2⟩)}.  The map defined by the
\leanline{=>}-notation maps every $(x_i)_{i\in I}$ to a function of
\leanline{j}, whose type is the subtype of \leanline{I}, consisting of
a value and a proof of \leanline{J ⊆ I}. In other words, this is
$(x_j)_{j\in J}$.

@@

We note that we extend the standard assumption that all $\alpha_t$ are
separable, complete metric spaces (or Polish, i.e.\ separable and
metrizable through a complete metric) to cover the case of extended
pseude-metric spaces. Such spaces do not satisfy the frequently used
Hausdorff (or t2) property, i.e.\ there can be $x\neq y$ such that all
open balls around $x$ and $y$ overlap.
  This generalization was only possible since underlying results in
  \leanline{mathlib} were already provided on the same level of
  generality.
  More precisely, \leanline{isCompact_iff_totallyBounded_isComplete}, which shows that
  a set $A \subseteq \alpha$ is compact iff it is complete and totally
  bounded (see note~\ref{note:tot}), requires $\alpha$
  to be a uniform space (Remark~\ref{rem:uniform}, recall that every metric space is uniform).
  We also require the underlying space(s) to be second-countable (used in the proof of Lemma~\ref{L:relcoPol}).
  A second-countable uniform space can be made into an (extended) pseudo-metric space
  (\leanline{UniformSpace.pseudoMetrizableSpace}); see also
  Remark~\ref{rem:uniform} for some more details.

@@

Membership of that set is characterized as follows:
\begin{minted}[highlightlines={1}, framesep = 0pt, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem mem_cylinders
    (t : Set (∀ i : ι, α i)) :
    t ∈ cylinders α ↔
      ∃ (s S : _) (_ : MeasurableSet S), t = cylinder s S
\end{minted}
\noindent
The finite set \leanline{s} of the potential coordinates where
the cylinder deviates from the whole set needs not be unique.
From the exists-statement of \leanline{mem_cylinders}, we however choose an
arbitrary such \leanline{Finset ι}, called \leanlinecolor{cylinders.finset}.

@@

\begin{minted}[highlightlines={1-3}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem setField_cylinders : SetField (cylinders α) 
theorem setRing_cylinders :  SetRing (cylinders α)
theorem setSemiringCylinders : SetSemiring (cylinders α)   
\end{minted}

@@

Extending Lemma~\ref{l1}, we would like to write a finite union of
members of a semi-ring as a finite union of disjoint sets\todo{Is this section going into too much detail?}.

@@

\noindent
The formalized version again gives rise to a definition:

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def indexedDiff₀ (hC : SetSemiring C)
    (J : Finset (Set α)) (hJ : ↑J ⊆ C)
    (n : Fin J.card) : Finset (Set α) :=
  hC.diff₀ (ordered_mem' hJ n) (finsetLT J n)
    (finsetLT_subset' J hJ n)
\end{minted}

Here, recall that \leanline{Fin n} is the subtype of \leanline{Nat}
consisting of all numbers \leanline{<n}. In addition,
\leanline{ordered_mem' (hJ : ↑J ⊆ C) (n : Fin (card J)) : ↑(ordered J)
  n ∈ C} gives a proof that the \leanline{n}th element of \leanline{J}
is in \leanline{C}, \leanline{finsetLT (J : Finset (Set α)) (n : Fin (card J)) : Finset (Set α)} is the set consisting of the
\leanline{n} sets in \leanline{J} numbered $0,...,n-1$, and
\leanlinecolor{finsetLT_subset'} gives a proof that
\leanline{↑(finsetLT J n) ⊆ C}.  So, in terms of Lemma~\ref{l2}, this
gives, for $n=1,...,m$, a construction for $\mathcal K_n \subseteq_f
\mathcal H$.

\noindent
\subsection{Carathéodory's extension theorem}
Measures are set functions defined on a $\sigma$-algebra (i.e.\ an
algebra stable under countable unions), satisfying some properties
which we recall below. Mostly, defining such a measure on the full
$\sigma$-algebra is not possible, but defining a set function on a
semi-ring is possible. As an example, consider Lebesgue-measure on
$\mathbb R$, with the Borel $\sigma$-algebra on $\mathbb R$, which is
$\sigma(\mathcal O)$, where $\mathcal O$ is the set of open sets, and
$\sigma(\mathcal O)$ is the smallest $\sigma$-algebra containing all
open sets. Since this set system is defined only abstractly, it is
hard to know which volume each of these sets should be assigned to at
first sight. However, the volume of an interval is easy, since it may
be defined by the length of the interval. So, in order to construct
measures from a set function $m$ on a (semi-)ring $\mathcal H$
(e.g.\ the set of all semi-open intervals), it has been a fundamental
insight of Carathéodory that one may start by defining a set function
(outer measure) $\mu$ on $2^\alpha$, and then show $\mu$ extends $m$
and defines a measure on $\sigma(\mathcal H)$. (Note that the set of
semi-open intervals generates the Borel $\sigma$-algebra on $\mathbb
R$.) We will follow this abstract construction, and start by stating
some basic concepts.

@@

\noindent
For additive contents, we need two definitions.
The structure \leanlinecolor{AddContent} defines
an additive set-function \leanline{toFun} on \leanline{Set α}.
The definition \leanlinecolor{extendContent} takes an additive function defined only on a semi-ring of sets, and extends it to an additive content on all sets by choosing the value infinity outside of the semi-ring. We use this because total functions are easier to use in subsequent code.

@@

Here, \leanlinecolor{kolmogorovFun_empty} and
\leanlinecolor{kolmogorovFun_additive} give proofs for
\leanline{m_empty} and \leanline{m_add} in
\leanlinecolor{extendContent} as applied to a
\leanlinecolor{kolmogorovFun}.

@@

\begin{theorem}[\boldmath $\mu$-measurable sets are a
    $\sigma$-algebra]\label{T:cara} Let $\mu$ be an outer measure on
  $E$ and $\mathcal F$ the set of $\mu$-measurable sets,
  i.e.\ the set of sets $A$ satisfying
  \begin{align}
    \label{eq:mumess}
    \mu(B) = \mu(B\cap A) + \mu(B\cap A^c), \qquad B \subseteq E.
  \end{align}
  Then, $\mathcal F$ is a $\sigma$-Algebra and $\mu|_{\mathcal F}$ is
  a measure. In addition, $\{N\subseteq \Omega:
  \mu(N)=0\}\subseteq \mathcal F$, i.e.\ $\mathcal F$ is complete.
\end{theorem}

@@

Here is a formalization of the measure resulting from
Theorem~\ref{T:masseind}:

@@

The above definition closely resembles its formalization in
\leanline{mathlib}:

@@

Here is the formalization:

@@

@@

Note that the assumption \leanline{hP_inner} from above translates
directly to inner regularity of the \leanlinecolor{kolContent}, which
is defined through the projective family \leanline{P}. Moreover, since
the Carathéodory extension theorem requires $\sigma$-subadditivity
(rather than $\sigma$-additivity), we can use
\leanlinecolor{sigma_subadditive_of_sigma_additive} in order
to show
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular}, and
the assumptions of the last result also imply $\sigma$-subadditivity.

@@

What remains to be done is to show conditions under which the
projective family is inner regular with respect to compact (and
closed) sets. For this, we need some assumption on the underlying
spaces. The following lemma provides the generalizations to extended
pseudo-metric spaces:

@@

In the next lemma, we will need that a closed subset of a compact set
is compact. (See \leanline{isCompact_of_isClosed_subset}.) We describe
the formalization of the next lemma only in conjunction with
Lemma~\ref{L:relcoPol}.

@@

\todo{too much detail? omit the proof?}

\begin{proof}
  2. $\Rightarrow$ 1.: This is clear since $\alpha \in \mathcal B$.
  1. $\Rightarrow$ 2.: Fix $\varepsilon>0$ and $A \in \mathcal B$. Let
  $K$ be closed and compact with $\mu(K^c) < \varepsilon/2$. By
  Lemma~\ref{l:pseude1}, there is some closed $F \subseteq A$
  with $\mu(A) < \mu(F) + \varepsilon/2$. Now, we have that $F \cap K
  \subseteq A$ is closed and compact, and with $A \setminus (F\cap K)
  = (A \setminus F) \cup (A \setminus K)$,
  $$\mu(A \setminus (F\cap K)) \leq \mu(A \setminus F) + \mu(A
  \setminus K) < \varepsilon/2 + \mu(K^c) < \varepsilon. $$
\end{proof}

@@

Let us combine the last two results in one combined formalized result.
This the gives the desired result that a measure on a second
countable, complete extended pseudo-metric space is inner regular with
respect to closed compact sets.

\begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem innerRegular_isCompact_isClosed_
    measurableSet_of_complete_countable
    [PseudoEMetricSpace α] [CompleteSpace α]
    [SecondCountableTopology α]
    [BorelSpace α] (P : Measure α) [IsFiniteMeasure P] :
    P.InnerRegular (fun s => IsCompact s ∧ IsClosed s)
      MeasurableSet 
\end{minted}

@@

\subsection{Proof of Kolmogorov's extension theorem}
We now describe the proof of Kolmogorov's extension theorem as well as
its formalization: For the proof of Theorem~\ref{T1}, we 
\begin{enumerate}
  \item apply Theorem~\ref{T:masseind} for the ring (hence semi-ring)
    $\mathcal A$ and the set-function $P$ as given in
    Theorem~\ref{T1};
  \item show $\sigma$-additivity (as assumed in
    Theorem~\ref{T:masseind}) of $P$ by using
    Lemma~\ref{l:stetigcompact}. For the latter, we need to show that
    $P$ is inner regular with respect to a compact system. Here, note
    that $\{\pi_J^{-1}C : C \in \prod_{j\in J} \alpha_j \text{ compact
      and closed}\}$ is a compact system and $P(\pi_J^{-1}C) =
    P_J(C)$, so we need to show that $P_J$ is inner regular with
    respect to compact and closed sets, $J\subseteq_f \iota$;
  \item use Lemma~\ref{L:relcoPol} in combination with
    Lemma~\ref{l:tight} (1.$\Rightarrow$2.) and the properties of
    completeness, separability of the underlying extended
    pseudo-metric spaces in order to see that every $P_J$ has the
    desired property of being inner regular wrt compact and closed
    sets. This concludes the proof of Theorem~\ref{T1}. 
\end{enumerate}
The formalization of this proof resembles these arguments. We leave
out all instances in the reformulation of the result and its proof
(see below Theorem~\ref{T1} for a full formulation):

@@

\section{Concluding thoughts}
\label{ss:concl}
{\color{blue} The usability of our contribution must be proved by an
  example: Adding to our general theory, we have implemented the
  example of an infinite product measure. (As an application of
  \leanlinecolor{projectiveLimitWithWeakestHypotheses}, we have to use
  complete and separable extended pseudo-metric spaces, but note that
  an implementation of the Ionescu-Tulcea Theorem would give
  the same result without these prerequisites.) Here, we can build on
  \leanline{Measure.pi}, which defines a finite product measure.
  Using the same automatically inferred instances, we obtain:}

>>>>

% Anonymous paper:
% \documentclass[sigplan,10pt,anonymous,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

% Paper with authors:
\documentclass[sigplan,10pt,review]{acmart}\settopmatter{printfolios=true,printccs=false,printacmref=false}

\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{ulem}
%\usepackage{amsmath}

@@

\usepackage{minted}
\setminted{encoding=utf-8}
\usepackage{fontspec}
%\setmainfont{dejavu-serif}
\setmonofont[Scale=0.833]{dejavu-sans}
%\setmainfont{FreeSerif}
%\setmonofont{FreeMono}
\usepackage{xcolor, graphics}
\definecolor{mygray}{rgb}{0.97,0.97,0.97}
\definecolor{LightCyan}{rgb}{0.8784,1,1}
\newcommand{\leanline}[1]{\mintinline[breaklines, breakafter=_., breaklines]{Lean}{#1}}%\textsuperscript{\textcolor{gray}{\tiny(m)}}}
\newcommand{\leanlinecolor}{\mintinline[breaklines, breakafter=_., breaklines, bgcolor=LightCyan]{Lean}}
%\usepackage{refcheck}

@@

\begin{abstract}
  We present a formalization of the Kolmogorov extension theorem in
  Lean. This result is a main building block in the construction of
  the (distribution of) stochastic processes with arbitrary index
  sets. Our approach is based on mathlib, the mathematical library for
  Lean. On our way, we provide a formalization of the Caratheodory
  extension theorem, which allows to extend a set function on a
  semi-ring to a proper measure on the $\sigma$-algebra generated by
  the semi-ring. A slight generalization of the classical Kolmogorov
  extension theorem allows us to construct the (distribution of) a
  stochastic process on a complete and separable pseudo-metric space
  rather than a metric space. {\color{blue}As an example, we define a
    product probability measure with arbitrary index set.}
\end{abstract}

@@

At first sight, it might be surprising that martingales, a certain class of stochastic process, have already been formalized in Lean \cite{ying2023formalization}, although the Kolmogorov extension theorem (on the existence of stochastic processes) is only available by our contribution. Note, however, that a martingale $(X_t)_{t\in \iota}$ is defined as a family of random variables (satisfying some properties) on a fixed probability space $(\Omega, \mathcal F, \mathbf P)$, while the Kolmogorov extension theorem is on the construction of a probability (or finite) measure $\mathbf P$, on which we can define random variables (with certain properties). So, our work complements \cite{ying2023formalization} in the sense that probability spaces $(\Omega, \mathcal F, \mathbf P)$ on which you can define a martingale exist. As an application of our implementation, we construct a probability measure on an infinite product space where all coordinates are independent; see Section~\ref{ss:concl}. {\color{blue} Our code is available at xxx.}

@@

\noindent
{\em Theorem of Ionescu-Tulcea and related results:}
While Kolmogorov's extension Theorem gives a result for arbitrary $\iota$, but requires some properties of the family $(\alpha_t)_{t\in\iota}$, the Theorem of Ionescu-Tulcea \cite[Theorem 8.24]{Kallenberg2020} can only deal with countable $\iota$, but has no restrictions on $(\alpha_t)_{t\in\iota}$. The proof of the latter, however, uses transition kernels (and an inductive construction of a content, which can be extended to a measure) rather than projective families.
However, the structure of the argument is similar and could use some of our results, like the proof of $\sigma$-additivity from continuity at the empty set (see Lemma~\ref{P:stetigmass}).
A formalization would complement our work. In addition, using the Theorem of Ionescu-Tulcea one can define infinitely many independent random variables (arbitrary $\iota$) on a joint probability space \cite[Theorem 8.24]{Kallenberg2020} {\color{blue} with fewer requirements on $(\alpha_t)_{t\in\iota}$ than in the example given in Section~\ref{ss:concl}}.
All of these results can be used in order to construct countably many independent random variables, which are frequently used in concrete constructions in probability theory. Examples are (one direction of) the Borel-Cantelli-Lemma \cite[Theorem 4.28]{Kallenberg2020}, Kolmogorov's 0-1-law \cite[Theorem 4.13]{Kallenberg2020}, random walks \cite[Chapter 12]{Kallenberg2020}, branching processes  \cite[Chapter 13]{Kallenberg2020}, and percolation  \cite[Chapter 2.4]{Klenke2013}, to name just a few. While the first two are already formalized in \leanline{mathlib}, it is only clear due to the Kolmorov extension theorem (or the Theorem of Ionesu-Tulcea) that infinitely many independent random variables can indeed be defined on a joint probability space. In this sense, our contribution is also important for giving more sense to already formalised parts of \leanline{mathlib}.

@@

We now come to our contribution. 
The extension theorem is a statement about extending a
set-function on a product space to a (finite) measure, where the
product space can come with an arbitraty index set.  The next
definition covers the important concept of a projective family of
measures. In short, we define measures on any finite subset of indices
in a consistent way. Finite sets (of some type \leanline{ι}) are
formalized with \leanline{Finset ι}. % Such a \leanline{Finset} is
defined as a \leanline{Multiset}.
%(a quotient type of the \leanline{List}-type with respect to
%permutations of the list) together with a proof that it contains no
%duplicates\todo{remove this implementation detail?}.

@@

In contrast to statements already part of \leanline{mathlib}, in the
sequel we \leanlinecolor{highlight} definitions and statements which
are part of our own code. In the formalization of the above, we use
\leanline{variable {ι : Type _} {α: ι → Type _}}, which fixes the
index set $\iota$ and all spaces $\alpha_t, t\in\iota$ as global
variables.  In Lean, the product space is then \leanline{∀ j : ι, α j}
and the product over a finite set \leanline{J} is \leanline{∀ j : J, α
  j}. {\color{blue} Note that \leanline{J} is a subtype of
  \leanline{ι} in this formulation.}

@@

\noindent
It is worth understanding the precise connection of
\leanlinecolor{isProjective} and
\leanlinecolor{isProjectiveMeasureFamily}. In the latter, the first
variable of \leanlinecolor{isProjective} is the family \leanline{P} of
finite measures for all finite subsets of \leanline{ι}. The second
variable is the functions which maps two sets \leanline{I J : Finset
  ι} and a proof \leanline{hJI} of \leanline{J ⊆ I} together with
\leanline{P I} (which is \leanline{μ} in the statement) to the image
measure on \leanline{J}, which is \leanline{μ.map (fun x : ∀ i : I, α
  i => fun j => x ⟨j, hJI j.2⟩)}.  The map defined by the
\leanline{=>}-notation maps every $(x_i)_{i\in I}$ to a function of
\leanline{j}, whose type is the subtype {\color{blue}\leanline{j} of}
\leanline{I}, consisting of a value and a proof of \leanline{J ⊆
  I}. In other words, this is $(x_j)_{j\in J}$.

@@

\noindent
Note that we extend the standard assumption that all $\alpha_t$ are
separable, complete metric spaces (or Polish, i.e.\ separable and
metrizable through a complete metric) to cover the case of extended
pseude-metric spaces. Such spaces do not satisfy the frequently used
Hausdorff (or t2) property, i.e.\ there can be $x\neq y$ such that all
open balls around $x$ and $y$ overlap.  This generalization was only
possible since underlying results in \leanline{mathlib} were already
provided on the same level of generality.  More precisely,
\leanline{isCompact_iff_totallyBounded_isComplete}, which shows that a
set $A \subseteq \alpha$ is compact iff it is complete and totally
bounded (see note~\ref{note:tot}), requires $\alpha$ to be a uniform
space (Remark~\ref{rem:uniform}, recall that every metric space is
uniform).  We also require the underlying space(s) to be
second-countable (used in the proof of Lemma~\ref{L:relcoPol}).  A
second-countable uniform space can be made into an (extended)
pseudo-metric space (\leanline{UniformSpace.pseudoMetrizableSpace});
see also Remark~\ref{rem:uniform} for some more details.

@@

\noindent
Membership of that set is characterized as follows:
\begin{minted}[highlightlines={1}, framesep = 0pt, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem mem_cylinders
    (t : Set (∀ i : ι, α i)) :
    t ∈ cylinders α ↔
      ∃ (s S : _) (_ : MeasurableSet S), t = cylinder s S
\end{minted}
\noindent
The finite set \leanline{s} of the potential coordinates where
the cylinder deviates from the whole set need not be unique.
From the $\exists$-statement of \leanline{mem_cylinders}, we however choose an
arbitrary such \leanline{Finset ι}, called \leanlinecolor{cylinders.finset}.

@@

\begin{minted}[highlightlines={1-4}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem setField_cylinders : SetField (cylinders α) 
theorem setRing_cylinders :  SetRing (cylinders α)
theorem setSemiringCylinders
  : SetSemiring (cylinders α)   
\end{minted}

@@

\noindent
Extending Lemma~\ref{l1}, we would like to write a finite union of
members of a semi-ring as a finite union of disjoint sets.

@@

%% \noindent
%% The formalized version again gives rise to a definition:

%% \begin{minted}[highlightlines={1}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
%% def indexedDiff₀ (hC : SetSemiring C)
%%     (J : Finset (Set α)) (hJ : ↑J ⊆ C)
%%     (n : Fin J.card) : Finset (Set α) :=
%%   hC.diff₀ (ordered_mem' hJ n) (finsetLT J n)
%%     (finsetLT_subset' J hJ n)
%% \end{minted}

%% Here, recall that \leanline{Fin n} is the subtype of \leanline{Nat}
%% consisting of all numbers \leanline{<n}. In addition,
%% \leanline{ordered_mem' (hJ : ↑J ⊆ C) (n : Fin (card J)) : ↑(ordered J)
%%   n ∈ C} gives a proof that the \leanline{n}th element of \leanline{J}
%% is in \leanline{C}, \leanline{finsetLT (J : Finset (Set α)) (n : Fin (card J)) : Finset (Set α)} is the set consisting of the
%% \leanline{n} sets in \leanline{J} numbered $0,...,n-1$, and
%% \leanlinecolor{finsetLT_subset'} gives a proof that
%% \leanline{↑(finsetLT J n) ⊆ C}.  So, in terms of Lemma~\ref{l2}, this
%% gives, for $n=1,...,m$, a construction for $\mathcal K_n \subseteq_f
%% \mathcal H$.

\noindent
\subsection{Carathéodory's extension theorem}
Measures are set functions defined on a $\sigma$-algebra, satisfying
some properties which we recall below. {\color{blue}Frequently,
  constructively} defining such a measure on the full $\sigma$-algebra
is not possible, but defining a set function on a semi-ring is
possible. As an example, consider Lebesgue-measure on $\mathbb R$,
with the Borel $\sigma$-algebra on $\mathbb R$, which is
$\sigma(\mathcal O)$, where $\mathcal O$ is the set of open sets, and
$\sigma(\mathcal O)$ is the smallest $\sigma$-algebra containing all
open sets. Since this set system is defined only abstractly, it is
hard to know which volume each of these sets should be assigned to at
first sight. However, the volume of an interval is easy, since it may
be defined by the length of the interval. So, in order to construct
measures from a set function $m$ on a (semi-)ring $\mathcal H$
(e.g.\ the set of all semi-open intervals), it has been a fundamental
insight of Carathéodory that one may start by defining a set function
(outer measure) $\mu$ on $2^\alpha$, and then show $\mu$ extends $m$
and defines a measure on {\color{blue}the $\sigma$-algebra of
  measurable sets (see Theorem~\ref{T:cara} below), which contains}
$\sigma(\mathcal H)$.
%(Note that the set of semi-open intervals
%generates the Borel $\sigma$-algebra on $\mathbb R$.)
We will follow this abstract construction, and start by stating some
basic concepts.

@@

\noindent
For additive contents, we need two definitions.
The structure \leanlinecolor{AddContent} defines
an additive set-function on \leanline{Set α}.
The definition \leanlinecolor{extendContent} takes an additive function defined only on a semi-ring of sets, and extends it to an additive content on all sets by choosing the value infinity outside of the semi-ring. We use this because total functions are easier to use in subsequent code.

@@

Here, \leanlinecolor{kolmogorovFun_empty} and
\leanlinecolor{kolmogorovFun_⌋} \leanlinecolor{additive} give proofs for
\leanline{m_empty} and \leanline{m_add} in
\leanlinecolor{extendContent} as applied to a
\leanlinecolor{kolmogorovFun}.

@@

\begin{theorem}[\boldmath $\mu$-measurable sets are a
    $\sigma$-algebra]\label{T:cara} Let $\mu$ be an outer measure on
  $E$ and $\mathcal F$ the set of $\mu$-measurable sets,
  i.e.\ the set of sets $A$ satisfying
  \begin{align*}
    \mu(B) = \mu(B\cap A) + \mu(B\cap A^c), \qquad B \subseteq E.
  \end{align*}
  Then, $\mathcal F$ is a $\sigma$-Algebra and $\mu|_{\mathcal F}$ is
  a measure. In addition, $\{N\subseteq \Omega:
  \mu(N)=0\}\subseteq \mathcal F$, i.e.\ $\mathcal F$ is complete.
\end{theorem}

@@

\noindent
Here is a formalization of the measure resulting from
Theorem~\ref{T:masseind}:

@@

\noindent
The above definition closely resembles its formalization in
\leanline{mathlib}:

@@

\noindent
Here is the formalization:

@@

\noindent
{\color{blue} Such compact systems are important since they allow a
  proof of $\sigma$-additivity of a content on a ring, which is the
  missing piece for applying the Carathéodory Theorem in the proof of
  the Kolmogorov extension theorem.}

@@

Note that the assumption \leanline{hP_inner} from above translates
directly to inner regularity of the \leanlinecolor{kolContent}, which
is defined through the projective family \leanline{P}. Moreover, since
the Carathéodory extension theorem requires $\sigma$-subadditivity
(rather than $\sigma$-additivity), we can use
\leanlinecolor{sigma_subadditive_of_sigma_additive} {\color{blue}(see
  Lemma~\ref{P:stetigmass}, 1.$\Rightarrow$2.)} in order to show that
the assumptions of the last result also imply $\sigma$-subadditivity;
this gives
\leanlinecolor{kolContent_sigma_subadditive_of_innerRegular}.

@@

What remains to be done is to show conditions under which the
projective family is inner regular with respect to compact (and
closed) sets. For this, we need some assumption on the underlying
spaces.

@@

In the next lemma, we will need that a closed subset of a compact set
is compact. (See \leanline{isCompact_of_isClosed_subset}.) We describe
the formalization of the next lemma only in conjunction with
Lemma~\ref{L:relcoPol}. {\color{blue} We omit the proof, since the
  only interesting part is an application of inner regularity from
  Lemma~\ref{l:pseude1}.}

@@

%% \todo{too much detail? omit the proof?}

%% \begin{proof}
%%   2. $\Rightarrow$ 1.: This is clear since $\alpha \in \mathcal B$.
%%   1. $\Rightarrow$ 2.: Fix $\varepsilon>0$ and $A \in \mathcal B$. Let
%%   $K$ be closed and compact with $\mu(K^c) < \varepsilon/2$. By
%%   Lemma~\ref{l:pseude1}, there is some closed $F \subseteq A$
%%   with $\mu(A) < \mu(F) + \varepsilon/2$. Now, we have that $F \cap K
%%   \subseteq A$ is closed and compact, and with $A \setminus (F\cap K)
%%   = (A \setminus F) \cup (A \setminus K)$,
%%   $$\mu(A \setminus (F\cap K)) \leq \mu(A \setminus F) + \mu(A
%%   \setminus K) < \varepsilon/2 + \mu(K^c) < \varepsilon. $$
%% \end{proof}

@@

Let us combine the last two results in one formalized result.  This
the gives the desired statement that a measure on a second countable,
complete extended pseudo-metric space is inner regular with respect to
closed compact sets.

\begin{minted}[highlightlines={1-2}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
theorem innerRegular_isCompact_isClosed_
    measurableSet_of_complete_countable
    [PseudoEMetricSpace α] [CompleteSpace α]
    [SecondCountableTopology α]
    [BorelSpace α] (P : Measure α) [IsFiniteMeasure P] :
    P.InnerRegular (fun s => IsCompact s ∧ IsClosed s)
      MeasurableSet 
\end{minted}

@@

\subsection{Proof of Kolmogorov's extension theorem}
We now describe the proof of Kolmogorov's extension theorem as well as
its formalization: For the proof of Theorem~\ref{T1}, we 
\begin{enumerate}
  \item apply Theorem~\ref{T:masseind} for the ring (hence semi-ring)
    $\mathcal A$ and the set-function $P$ as given in
    Theorem~\ref{T1};
  \item show $\sigma$-additivity (as assumed in
    Theorem~\ref{T:masseind}) of $P$ by using
    Lemma~\ref{l:stetigcompact}. For the latter, we need to show that
    $P$ is inner regular with respect to a compact system. Here, note
    that $\{\pi_J^{-1}C : C \in \prod_{j\in J} \alpha_j \text{ compact
      and closed}\}$ is a compact system and $P(\pi_J^{-1}C) =
    P_J(C)$, so we need to show that $P_J$ is inner regular with
    respect to compact closed sets, $J\subseteq_f \iota$;
  \item use Lemma~\ref{L:relcoPol} in combination with
    Lemma~\ref{l:tight} (1.$\Rightarrow$2.) and the properties of
    completeness, separability of the underlying extended
    pseudo-metric spaces in order to see that every $P_J$ has the
    desired property of being inner regular wrt compact and closed
    sets. This concludes the proof of Theorem~\ref{T1}. 
\end{enumerate}
The formalization of this proof resembles these arguments. We leave
out all instances in the reformulation of the result and its proof
(see below Theorem~\ref{T1} for a full formulation):

@@

\section{Concluding thoughts}
\label{ss:concl}
The usability of our contribution must be proved by an example: Adding
to our general theory, we have implemented the example of an infinite
product measure.
%(As an application of
%\leanlinecolor{projectiveLimitWithWeakestHypotheses}, we have to
%use complete and separable extended pseudo-metric spaces, but note
%that an implementation of the Ionescu-Tulcea Theorem would give the
%same result without these prerequisites.)
Here, we can build on \leanline{Measure.pi}, which defines a finite
product measure.  Using the same automatically inferred instances,
we obtain:

\begin{minted}[highlightlines={1,8,13}, mathescape, numbersep=5pt, framesep=5mm, bgcolor=mygray]{Lean}
def independentFamily
    (P : ∀ i, Measure (α i)) [∀ i,
    IsProbabilityMeasure (P i)] : Measure (∀ i, α i) :=
  projectiveLimitWithWeakestHypotheses
  (fun J : Finset ι =>
  Measure.subset_pi P J) (product_isProjective P)
